---
title: "Arrest & Incarceration (adm)_by geo categories in incarceration data"
---


```{r setup, include=FALSE}
# Load packages
require(dplyr)
require(haven)
require(readr)
```

  
  
# Read and merge incarceration and arrest data 
*This chunk does not have to be run if the code is not changed. The data obtained in this chunk have been saved and can be loaded for running the the other chunks.*
```{r}
# Read incarceration data
setwd("O:/Arrest Trends - Phase II/Suburb Analysis/Data")
jail <- read_csv("incarceration_trends.csv")

## Subset to years after 1977
jail <- jail[jail$year >= 1977, ] %>%
  rename(YEAR = year, FIPS = fips)

# Read county level estimated arrest data
#setwd("O:/Arrest Trends - Phase II/Final Data CSV/Estimated Datasets/County Level")
arrest <- read_csv("Final_County_SES_Arrests.csv")

# Merge the two datasets
jail$FIPS <- sprintf("%06d", jail$FIPS) %>%
  substr(2, 6)
jail_arrest <- full_join(arrest, jail, by = c("FIPS", "YEAR"))

# Pick the variables needed
jail_arrest <- jail_arrest %>% 
  select(FIPS, YEAR, COUNTYNAME, FSTATE, REGION,
         total_jail = total_jail_adm, 
         total_arrest = TOTAL_COUNTY_ARRESTS, 
         rate_arrest = TOTAL_COUNTY_ARRESTRATE, 
         pop_jail = total_pop,
         pop_arrest = POP,
         urbanicity,
         UnemploymentRate,
         TOTAL_County_Uninsured,
         Pct_Uninsured,
         TOTAL_County_Poverty,
         Pct_Poverty,
         Median_Household_Inc)

# Convert -99999 to NA
jail_arrest$total_arrest[jail_arrest$total_arrest==-99999] <- NA
jail_arrest$rate_arrest[jail_arrest$rate_arrest==-99999] <- NA
jail_arrest$pop_arrest[jail_arrest$pop_arrest==-99999] <- NA


# Save the data
setwd("O:/Arrest Trends - Phase II/Suburb Analysis/Data")
save(jail_arrest, file = "jail_arrest_county.RData")
```



# Subset data to the four geographical types according to the variable "urbanicity"
```{r}
#setwd("O:/Arrest Trends - Phase II/Suburb Analysis/Data")
setwd("D:/NYU/CAREER/Vera Institute of Justice/Work Materials")
load("jail_arrest_county.RData")

# Rural
rural <- jail_arrest[jail_arrest$urbanicity == "rural", ] %>%
  filter(!is.na(YEAR)) %>%
  as.data.frame()

# Small / Mid
sm <- jail_arrest[jail_arrest$urbanicity == "small/mid", ] %>%
  filter(!is.na(YEAR)) %>%
  as.data.frame()

# Suburban
suburb <- jail_arrest[jail_arrest$urbanicity == "suburban", ] %>%
  filter(!is.na(YEAR)) %>%
  as.data.frame()

# Urban
urban <- jail_arrest[jail_arrest$urbanicity == "urban", ] %>%
  filter(!is.na(YEAR)) %>%
  as.data.frame()


urbanicity <- list(rural, sm, suburb, urban)
```



# Calculate rates

## Source the rate calculation function
```{r}
setwd("D:/NYU/CAREER/Vera Institute of Justice/Work Materials/Analysis-by-Geography/Scripts")
source("Rate Calculation Function.R")
```


# Calculate the rates for each geographical type
```{r}
rates <- list()
for (i in 1:4) {
  rates[[i]] <- cal_rates(urbanicity[[i]])
}

# In 1977 all the jail admission rates are 0, which does not look right.
# In 1978-1980, 2015-2016, either arrest data or incarceration data are missing.
# Remove these years
for (i in 1:4) {
  rates[[i]] <- rates[[i]] %>%
    filter(! (YEAR %in% c(1977:1980, 2015:2016)))
}
```


# Get the rates table
```{r}
# Arrest and jail admission rates by the three methods for the four geography types
rates_list <- list()
geo_names <- c("rural", "small/mid", "suburb", "urban")
for (i in 1:4) {
  dirc <- rates_list[[i]] <- cal_rates(urbanicity[[i]], method = "direct")
  rates_list[[i]]$arrest_dirc <- dirc$rate_arrest
  rates_list[[i]]$jail_dirc <- dirc$rate_jail
  
  simp <- cal_rates(urbanicity[[i]], method = "simple")
  rates_list[[i]]$arrest_simp <- simp$rate_arrest
  rates_list[[i]]$jail_simp <- simp$rate_jail
  
  wt <- cal_rates(urbanicity[[i]], method = "weighted")
  rates_list[[i]]$arrest_wt <- wt$rate_arrest
  rates_list[[i]]$jail_wt <- wt$rate_jail
  
  rates_list[[i]] <- rates_list[[i]] %>%
    mutate(geography = geo_names[i]) %>%
    select(YEAR, geography,
           arrest_dirc, arrest_simp, arrest_wt, 
           jail_dirc, jail_simp, jail_wt)
                          
}


# Combine them into one data frame
arrest_jail_rates <- rates_list[[1]]
for (i in 2:4) {
  arrest_jail_rates <- rbind(arrest_jail_rates, rates_list[[i]])
}
arrest_jail_rates <- arrest_jail_rates %>%
  arrange(YEAR, geography)

# Save the table in csv
setwd("D:/NYU/CAREER/Vera Institute of Justice/Work Materials/Analysis-by-Geography/Tables")
write.csv(arrest_jail_rates, file = "Arrest & Jail Adm Rates_All Data.csv", row.names = FALSE)
```


## Plot the incarceration trends
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_jail, type = "l", col = color[1],
     main = "Jail Admission Rate for Different Geographical Types", 
     xlab = "Year", ylab = "Rate", font.lab = 4,
     ylim = c(min(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail),
              max(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_jail, type = "l", col = color[i])
}

legend("topleft", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```

### Plot for blog post
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_jail, type = "l", col = color[1],
     main = "Jail Admission Rate by Geography", 
     xlab = "Year", ylab = "Rate", xaxt='n', yaxt='n', font.lab = 4, bty = 'n',
     ylim = c(0, max(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_jail, type = "l", col = color[i])
}

axis(side = 1, at = seq(from = 1980, to = 2015, by = 5), 
     labels = seq(from = 1980, to = 2015, by = 5), col = "azure3")
axis(side = 2, at = seq(from = 0, to = 6000, by = 1000), 
     labels = seq(from = 0, to = 6000, by = 1000), col = "azure3")

legend("bottomright", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```



## Plot the arrest trends
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_arrest, type = "l", col = color[1],
     main = "Arrest Rate for Different Geographical Types", 
     xlab = "Year", ylab = "Rate", font.lab = 4,
     ylim = c(min(rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE),
              max(rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_arrest, type = "l", col = color[i])
}

legend("topright", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```

### Plot for blog post
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_arrest, type = "l", col = color[1],
     main = "Arrest Rate by Geography", 
     xlab = "Year", ylab = "Rate", xaxt='n', yaxt='n', font.lab = 4, bty = 'n',
     ylim = c(0, max(rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_arrest, type = "l", col = color[i])
}

axis(side = 1, at = seq(from = 1980, to = 2015, by = 5), 
     labels = seq(from = 1980, to = 2015, by = 5), col = "azure3")
axis(side = 2, at = seq(from = 0, to = 6000, by = 1000), 
     labels = seq(from = 0, to = 6000, by = 1000), col = "azure3")

legend("bottomright", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```


# Plot the arrest and jail admission rates together
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

par(mar = c(5, 4, 4, 6) + 0.1, xpd = TRUE)
plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_jail, type = "l", col = color[1],
     main = "Jail Admission and Arrest Rate for Different Geographical Types", 
     xlab = "Year", ylab = "Arrest Rate", font.lab = 4,
     ylim = c(min(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail, rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE),
              max(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail, rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_jail, type = "l", col = color[i])
}
for (i in 1:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_arrest, type = "l", col = color[i], lty = "dashed")
}

legend("topright", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3, inset = c(-0.28, 0),
       legend = c("rural", "small/mid", "suburb", "urban"))

par(mar = c(5, 4, 4, 2) + 0.1)
```

### Plot the above image for blog post
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

par(mar = c(5, 4.5, 4, 4.5) + 0.1, xpd = TRUE)
plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_jail, type = "l", col = color[1],
     main = "Jail Admission and Arrest by Geography", 
     xlab = "Year", ylab = "Rate", xaxt='n', yaxt='n', font.lab = 4, bty = 'n',
     ylim = c(0, max(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail, rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_jail, type = "l", col = color[i])
}
for (i in 1:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_arrest, type = "l", col = color[i], lty = "dashed")
}

axis(side = 1, at = seq(from = 1980, to = 2016, by = 4), 
     labels = seq(from = 1980, to = 2016, by = 4), col = "azure3")
axis(side = 2, at = seq(from = 0, to = 6000, by = 1000), 
     labels = seq(from = 0, to = 6000, by = 1000), col = "azure3")

legend("topright", col = color, pch = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3, inset = c(-0.18, 0),
       legend = c("rural", "small/mid", "suburb", "urban"))
legend("bottomright", col = "black", lty = c("solid", "dashed"), bty = "n", text.width = 5, cex = 1, text.font = 3, inset = c(-0.06, 0),
       legend = c("Jail admission rate", "Arrest rate"))

par(mar = c(5, 4, 4, 2) + 0.1, xpd = TRUE)
```



# SES of counties in different geographical types
```{r}
# Write a function to calculate the weighted mean of SES 
cal_ses <- function(dat, sesvar, use.pop.jail = TRUE) {

  dat <- data.frame(dat, ses.var = sesvar) %>%
    arrange(YEAR)
  
  if (use.pop.jail) {
    dat <- dat %>%
      group_by(YEAR) %>%
      summarise(Total_POP = sum(pop_jail, na.rm = TRUE), 
                ses_wt = sum(ses.var*pop_jail, na.rm = TRUE),
                ## Creat indicators for whether all the data are missing for the year
                missing_ses = mean(is.na(ses.var))) %>%
      mutate(ses = ses_wt/Total_POP) %>%
      select(-ses_wt)
    
    # If all the data are missing for the year, then convert the 0 to NA
    dat$ses[dat$missing_ses==1] <- NA

  } else {
    dat <- dat %>%
      group_by(YEAR) %>%
      summarise(Total_POP = sum(pop_arrest, na.rm = TRUE), 
                ses_wt = sum(ses.var*pop_arrest, na.rm = TRUE),
                ## Creat indicators for whether all the data are missing for the year
                missing_ses = mean(is.na(ses.var))) %>%
      mutate(ses = ses_wt/Total_POP) %>%
      select(-ses_wt) 
    
    # If all the data are missing for the year, then convert the 0 to NA
    dat$ses[dat$missing_ses==1] <- NA
  }
  
  return(dat)
}
```

## Median income
```{r}
inc <- list()
for (i in 1:4) {
  inc[[i]] <- cal_ses(urbanicity[[i]], urbanicity[[i]]$Median_Household_Inc) %>%
    #Keep 1995-2014 when the data are annually consecutive
    filter(YEAR >= 1995, YEAR <=  2014)
}
```

### Plot the income trends
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = inc[[1]]$YEAR, y = inc[[1]]$ses, type = "l", col = color[1],
     main = "Household Income for Different Geographical Types", 
     xlab = "Year", ylab = "Income", font.lab = 4,
     ylim = c(min(inc[[1]]$ses, inc[[2]]$ses, inc[[3]]$ses, inc[[4]]$ses, na.rm = TRUE),
              max(inc[[1]]$ses, inc[[2]]$ses, inc[[3]]$ses, inc[[4]]$ses, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = inc[[i]]$YEAR, y = inc[[i]]$ses, type = "l", col = color[i])
}

legend("topleft", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))



# Histograme fo median income for each type

```


## Poverty percentage
```{r}
pov <- list()
for (i in 1:4) {
  pov[[i]] <- cal_ses(urbanicity[[i]], urbanicity[[i]]$Pct_Poverty) %>%
    #Keep 1995-2014 when the data are annually consecutive
    filter(YEAR >= 1995, YEAR <=  2014)
}
```

### Plot the poverty trends
Note that there is no county level poverty data at 1996, so the plot starts at 1997 even there is data in 1005.
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = pov[[1]]$YEAR, y = pov[[1]]$ses, type = "l", col = color[1],
     main = "Percentage of People below Poverty for Different Geographical Types", 
     xlab = "Year", ylab = "Poverty %", font.lab = 4,
     ylim = c(min(pov[[1]]$ses, pov[[2]]$ses, pov[[3]]$ses, pov[[4]]$ses, na.rm = TRUE),
              max(pov[[1]]$ses, pov[[2]]$ses, pov[[3]]$ses, pov[[4]]$ses, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = pov[[i]]$YEAR, y = pov[[i]]$ses, type = "l", col = color[i])
}

legend("topleft", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```

#### Plot for blog post
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = pov[[1]]$YEAR, y = pov[[1]]$ses, type = "l", col = color[1],
     main = "Percentage of People below Poverty for Different Geographical Types", 
     xlab = "Year", ylab = "Poverty %", font.lab = 4, xaxt='n', yaxt='n', font.lab = 4, bty = 'n',
     ylim = c(min(pov[[1]]$ses, pov[[2]]$ses, pov[[3]]$ses, pov[[4]]$ses, na.rm = TRUE),
              max(pov[[1]]$ses, pov[[2]]$ses, pov[[3]]$ses, pov[[4]]$ses, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = pov[[i]]$YEAR, y = pov[[i]]$ses, type = "l", col = color[i])
}

axis(side = 1, at = seq(from = 1995, to = 2015, by = 5), 
     labels = seq(from = 1995, to = 2015, by = 5), col = "azure3")
axis(side = 2, at = seq(from = 7, to = 19, by = 2), 
     labels = seq(from = 7, to = 19, by = 2), col = "azure3")

legend(x = 1994, y = 21, col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```



## Unemployment rate
```{r}
unemp <- list()
for (i in 1:4) {
  unemp[[i]] <- cal_ses(urbanicity[[i]], urbanicity[[i]]$UnemploymentRate) %>%
    #Keep 1995-2014 when the data are annually consecutive
    filter(YEAR >= 1995, YEAR <=  2014)
}
```

### Plot the poverty trends
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = unemp[[1]]$YEAR, y = unemp[[1]]$ses, type = "l", col = color[1],
     main = "Unemployment Rate for Different Geographical Types", 
     xlab = "Year", ylab = "Poverty %", font.lab = 4,
     ylim = c(min(unemp[[1]]$ses, unemp[[2]]$ses, unemp[[3]]$ses, unemp[[4]]$ses, na.rm = TRUE),
              max(unemp[[1]]$ses, unemp[[2]]$ses, unemp[[3]]$ses, unemp[[4]]$ses, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = unemp[[i]]$YEAR, y = unemp[[i]]$ses, type = "l", col = color[i])
}

legend("topleft", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```