---
title: "Arrest & Incarceration (adm)_by geo categories in incarceration data"
---


```{r setup, include=FALSE}
# Load packages
require(dplyr)
require(haven)
require(readr)
```

  
# Deal with missing data
```{r}
#setwd("O:/Arrest Trends - Phase II/Suburb Analysis/Data")
setwd("D:/NYU/CAREER/Vera Institute of Justice/Work Materials")
load("jail_arrest_county.RData")


# Get the FSTATE from the first two character of FIPS 
# because the original FSTATE is missing for some counties
jail_arrest$FSTATE <- as.numeric(substr(jail_arrest$FIPS, 1, 2))


# Calculate state totals of arrests for each year
state_arrest <- jail_arrest %>%
  group_by(FSTATE, YEAR) %>%
  summarise(total_arrest = sum(total_arrest, na.rm = T))
# In some years and some states, the total number of arrests is 0, which is unreasonable.
# We assume that these 0's are actually missing values.

# Get the FSTATE and YEAR with total 0 arrest
state_year_arrest0 <- paste(state_arrest$FSTATE[state_arrest$total_arrest==0], 
                            state_arrest$YEAR[state_arrest$total_arrest==0])

# Convert the arrest numbers and rates for all the counties in these states and years to NA
state_year <- paste(jail_arrest$FSTATE, jail_arrest$YEAR)
jail_arrest$total_arrest[state_year %in% state_year_arrest0] <- NA
jail_arrest$rate_arrest[state_year %in% state_year_arrest0] <- NA


# Repeat the steps above for jail admission
state_jail <- jail_arrest %>%
  group_by(FSTATE, YEAR) %>%
  summarise(total_jail = sum(total_jail, na.rm = T))
state_year_jail0 <- paste(state_jail$FSTATE[state_jail$total_jail==0], 
                          state_jail$YEAR[state_jail$total_jail==0])
#jail0 <- jail_arrest[state_year %in% state_year_jail0, ]
jail_arrest$total_jail[state_year %in% state_year_jail0] <- NA


# Keep only records with both non-missing arrest data and jail data, except 1993.
# In 1993 all the arrest data are missing while there jail admission data are not missing.
# So te records in 1993 are kept to calculate the jail admission rate.
jail_arrest <- jail_arrest %>%
  filter((!is.na(total_jail) & !is.na(total_arrest)) | YEAR == 1993) %>%
  # Filter out counties with missing jail admission data, no matter whether it is in 1993
  filter(!is.na(total_jail))
```


# Subset data to the four geographical types according to the variable "urbanicity"
```{r}
# Rural
rural <- jail_arrest[jail_arrest$urbanicity == "rural", ] %>%
  filter(!is.na(YEAR)) %>%
  as.data.frame()

# Small / Mid
sm <- jail_arrest[jail_arrest$urbanicity == "small/mid", ] %>%
  filter(!is.na(YEAR)) %>%
  as.data.frame()

# Suburban
suburb <- jail_arrest[jail_arrest$urbanicity == "suburban", ] %>%
  filter(!is.na(YEAR)) %>%
  as.data.frame()

# Urban
urban <- jail_arrest[jail_arrest$urbanicity == "urban", ] %>%
  filter(!is.na(YEAR)) %>%
  as.data.frame()


urbanicity <- list(rural, sm, suburb, urban)
```



# Calculate rates

## Source the rate calculation function
```{r}
setwd("D:/NYU/CAREER/Vera Institute of Justice/Work Materials/Analysis-by-Geography/Scripts")
source("Rate Calculation Function.R")
```


# Calculate the rates for each geographical type
```{r}
rates <- list()
for (i in 1:4) {
  rates[[i]] <- cal_rates(urbanicity[[i]])
}

# In 1977 all the jail admission rates are 0, which does not look right.
# In 1978-1980, 2015-2016, either arrest data or incarceration data are missing.
# Remove these years
for (i in 1:4) {
  rates[[i]] <- rates[[i]] %>%
    filter(! (YEAR %in% c(1977:1980, 2015:2016)))
}
```


# Get the rates table
```{r}
# Arrest and jail admission rates by the three methods for the four geography types
rates_list <- list()
geo_names <- c("rural", "small/mid", "suburb", "urban")
for (i in 1:4) {
  dirc <- rates_list[[i]] <- cal_rates(urbanicity[[i]], method = "direct")
  rates_list[[i]]$arrest_dirc <- dirc$rate_arrest
  rates_list[[i]]$jail_dirc <- dirc$rate_jail
  
  simp <- cal_rates(urbanicity[[i]], method = "simple")
  rates_list[[i]]$arrest_simp <- simp$rate_arrest
  rates_list[[i]]$jail_simp <- simp$rate_jail
  
  wt <- cal_rates(urbanicity[[i]], method = "weighted")
  rates_list[[i]]$arrest_wt <- wt$rate_arrest
  rates_list[[i]]$jail_wt <- wt$rate_jail
  
  rates_list[[i]] <- rates_list[[i]] %>%
    mutate(geography = geo_names[i]) %>%
    select(YEAR, geography,
           arrest_dirc, arrest_simp, arrest_wt, 
           jail_dirc, jail_simp, jail_wt)
                          
}


# Combine them into one data frame
arrest_jail_rates <- rates_list[[1]]
for (i in 2:4) {
  arrest_jail_rates <- rbind(arrest_jail_rates, rates_list[[i]])
}
arrest_jail_rates <- arrest_jail_rates %>%
  arrange(YEAR, geography)

# Save the table in csv
setwd("D:/NYU/CAREER/Vera Institute of Justice/Work Materials/Analysis-by-Geography/Tables")
write.csv(arrest_jail_rates, file = "Arrest & Jail Adm Rates_Missing data filtered.csv", row.names = FALSE)
```


## Plot the incarceration trends
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_jail, type = "l", col = color[1],
     main = "Jail Admission Rate for Different Geographical Types", 
     xlab = "Year", ylab = "Rate", font.lab = 4,
     ylim = c(min(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail),
              max(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_jail, type = "l", col = color[i])
}

legend("topleft", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```

### Plot for blog post
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_jail, type = "l", col = color[1],
     main = "Jail Admission Rate by Geography", 
     xlab = "Year", ylab = "Rate", xaxt='n', yaxt='n', font.lab = 4, bty = 'n',
     ylim = c(0, max(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_jail, type = "l", col = color[i])
}

axis(side = 1, at = seq(from = 1980, to = 2015, by = 5), 
     labels = seq(from = 1980, to = 2015, by = 5), col = "azure3")
axis(side = 2, at = seq(from = 0, to = 6000, by = 1000), 
     labels = seq(from = 0, to = 6000, by = 1000), col = "azure3")

legend("bottomright", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```



## Plot the arrest trends
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_arrest, type = "l", col = color[1],
     main = "Arrest Rate for Different Geographical Types", 
     xlab = "Year", ylab = "Rate", font.lab = 4,
     ylim = c(min(rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE),
              max(rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_arrest, type = "l", col = color[i])
}

legend("topright", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```

### Plot for blog post
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_arrest, type = "l", col = color[1],
     main = "Arrest Rate by Geography", 
     xlab = "Year", ylab = "Rate", xaxt='n', yaxt='n', font.lab = 4, bty = 'n',
     ylim = c(0, max(rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_arrest, type = "l", col = color[i])
}

axis(side = 1, at = seq(from = 1980, to = 2015, by = 5), 
     labels = seq(from = 1980, to = 2015, by = 5), col = "azure3")
axis(side = 2, at = seq(from = 0, to = 6000, by = 1000), 
     labels = seq(from = 0, to = 6000, by = 1000), col = "azure3")

legend("bottomright", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```


# Plot the arrest rate and jail admission rate together
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

par(mar = c(5, 4, 4, 6) + 0.1, xpd = TRUE)
plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_jail, type = "l", col = color[1],
     main = "Jail Admission and Arrest Rate for Different Geographical Types", 
     xlab = "Year", ylab = "Rate", font.lab = 4,
     ylim = c(min(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail, rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE),
              max(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail, rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_jail, type = "l", col = color[i])
}
for (i in 1:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_arrest, type = "l", col = color[i], lty = "dashed")
}

legend("topright", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3, inset = c(-0.28, 0),
       legend = c("rural", "small/mid", "suburb", "urban"))

par(mar = c(5, 4, 4, 2) + 0.1)
```

### Plot the above image for blog post
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

par(mar = c(5, 4.5, 4, 4.5) + 0.1, xpd = TRUE)

plot(x = rates[[1]]$YEAR, y = rates[[1]]$rate_jail, type = "l", col = color[1],
     main = "Jail Admission and Arrest Rate by Geography", 
     xlab = "Year", ylab = "Rate", xaxt='n', yaxt='n', font.lab = 4, bty = 'n',
     ylim = c(0, max(rates[[1]]$rate_jail, rates[[2]]$rate_jail, rates[[3]]$rate_jail, rates[[4]]$rate_jail, rates[[1]]$rate_arrest, rates[[2]]$rate_arrest, rates[[3]]$rate_arrest, rates[[4]]$rate_arrest, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_jail, type = "l", col = color[i])
}
for (i in 1:4) {
  lines(x = rates[[i]]$YEAR, y = rates[[i]]$rate_arrest, type = "l", col = color[i], lty = "dashed")
}

axis(side = 1, at = seq(from = 1980, to = 2016, by = 4), 
     labels = seq(from = 1980, to = 2016, by = 4), col = "azure3")
axis(side = 2, at = seq(from = 0, to = 6000, by = 1000), 
     labels = seq(from = 0, to = 6000, by = 1000), col = "azure3")

legend("topright", col = color, pch = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3, inset = c(-0.18, 0),
       legend = c("rural", "small/mid", "suburb", "urban"))
legend("bottomright", col = "black", lty = c("solid", "dashed"), bty = "n", text.width = 5, cex = 1, text.font = 3, inset = c(-0.06, 0),
       legend = c("Jail admission rate", "Arrest rate"))

par(mar = c(5, 4, 4, 2) + 0.1, xpd = FALSE)
```



# SES of counties in different geographical types
```{r}
# Write a function to calculate the weighted mean of SES 
cal_ses <- function(dat, sesvar, use.pop.jail = TRUE) {

  dat <- data.frame(dat, ses.var = sesvar) %>%
    arrange(YEAR)
  
  if (use.pop.jail) {
    dat <- dat %>%
      group_by(YEAR) %>%
      summarise(Total_POP = sum(pop_jail, na.rm = TRUE), 
                ses_wt = sum(ses.var*pop_jail, na.rm = TRUE),
                ## Creat indicators for whether all the data are missing for the year
                missing_ses = mean(is.na(ses.var))) %>%
      mutate(ses = ses_wt/Total_POP) %>%
      select(-ses_wt)
    
    # If all the data are missing for the year, then convert the 0 to NA
    dat$ses[dat$missing_ses==1] <- NA

  } else {
    dat <- dat %>%
      group_by(YEAR) %>%
      summarise(Total_POP = sum(pop_arrest, na.rm = TRUE), 
                ses_wt = sum(ses.var*pop_arrest, na.rm = TRUE),
                ## Creat indicators for whether all the data are missing for the year
                missing_ses = mean(is.na(ses.var))) %>%
      mutate(ses = ses_wt/Total_POP) %>%
      select(-ses_wt) 
    
    # If all the data are missing for the year, then convert the 0 to NA
    dat$ses[dat$missing_ses==1] <- NA
  }
  
  return(dat)
}
```

## Median income
```{r}
inc <- list()
for (i in 1:4) {
  inc[[i]] <- cal_ses(urbanicity[[i]], urbanicity[[i]]$Median_Household_Inc) %>%
    #Keep 1995-2014 when the data are annually consecutive
    filter(YEAR >= 1995, YEAR <=  2014)
}
```

### Plot the income trends
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = inc[[1]]$YEAR, y = inc[[1]]$ses, type = "l", col = color[1],
     main = "Household Income for Different Geographical Types", 
     xlab = "Year", ylab = "Income", font.lab = 4,
     ylim = c(min(inc[[1]]$ses, inc[[2]]$ses, inc[[3]]$ses, inc[[4]]$ses, na.rm = TRUE),
              max(inc[[1]]$ses, inc[[2]]$ses, inc[[3]]$ses, inc[[4]]$ses, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = inc[[i]]$YEAR, y = inc[[i]]$ses, type = "l", col = color[i])
}

legend("topleft", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))



# Histograme fo median income for each type

```


## Poverty percentage
```{r}
pov <- list()
for (i in 1:4) {
  pov[[i]] <- cal_ses(urbanicity[[i]], urbanicity[[i]]$Pct_Poverty) %>%
    #Keep 1995-2014 when the data are annually consecutive
    filter(YEAR >= 1995, YEAR <=  2014)
}
```

### Plot the poverty trends
Note that there is no county level poverty data at 1996, so the plot starts at 1997 even there is data in 1005.
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = pov[[1]]$YEAR, y = pov[[1]]$ses, type = "l", col = color[1],
     main = "Percentage of People below Poverty for Different Geographical Types", 
     xlab = "Year", ylab = "Poverty %", font.lab = 4,
     ylim = c(min(pov[[1]]$ses, pov[[2]]$ses, pov[[3]]$ses, pov[[4]]$ses, na.rm = TRUE),
              max(pov[[1]]$ses, pov[[2]]$ses, pov[[3]]$ses, pov[[4]]$ses, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = pov[[i]]$YEAR, y = pov[[i]]$ses, type = "l", col = color[i])
}

legend("topleft", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```

#### Plot for blog post
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = pov[[1]]$YEAR, y = pov[[1]]$ses, type = "l", col = color[1],
     main = "Percentage of People below Poverty for Different Geographical Types", 
     xlab = "Year", ylab = "Poverty %", font.lab = 4, xaxt='n', yaxt='n', font.lab = 4, bty = 'n',
     ylim = c(min(pov[[1]]$ses, pov[[2]]$ses, pov[[3]]$ses, pov[[4]]$ses, na.rm = TRUE),
              max(pov[[1]]$ses, pov[[2]]$ses, pov[[3]]$ses, pov[[4]]$ses, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = pov[[i]]$YEAR, y = pov[[i]]$ses, type = "l", col = color[i])
}

axis(side = 1, at = seq(from = 1995, to = 2015, by = 5), 
     labels = seq(from = 1995, to = 2015, by = 5), col = "azure3")
axis(side = 2, at = seq(from = 7, to = 19, by = 2), 
     labels = seq(from = 7, to = 19, by = 2), col = "azure3")

legend(x = 1994, y = 21, col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```



## Unemployment rate
```{r}
unemp <- list()
for (i in 1:4) {
  unemp[[i]] <- cal_ses(urbanicity[[i]], urbanicity[[i]]$UnemploymentRate) %>%
    #Keep 1995-2014 when the data are annually consecutive
    filter(YEAR >= 1995, YEAR <=  2014)
}
```

### Plot the poverty trends
```{r}
color = c("gray30", "coral4", "chartreuse4", "blue4")

#par(mar = c(5, 4, 6, 2) + 0.1, xpd = TRUE)
plot(x = unemp[[1]]$YEAR, y = unemp[[1]]$ses, type = "l", col = color[1],
     main = "Unemployment Rate for Different Geographical Types", 
     xlab = "Year", ylab = "Poverty %", font.lab = 4,
     ylim = c(min(unemp[[1]]$ses, unemp[[2]]$ses, unemp[[3]]$ses, unemp[[4]]$ses, na.rm = TRUE),
              max(unemp[[1]]$ses, unemp[[2]]$ses, unemp[[3]]$ses, unemp[[4]]$ses, na.rm = TRUE)))
for (i in 2:4) {
  lines(x = unemp[[i]]$YEAR, y = unemp[[i]]$ses, type = "l", col = color[i])
}

legend("topleft", col = color, lty = 1, bty = "n", text.width = 5, cex = 1.09, text.font = 3,
       legend = c("rural", "small/mid", "suburban", "urban"))
```